\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T2A]{fontenc}
\usepackage[mytitle={Oleksandr Kulkov. ETH Zürich. Linear Algebra. Week 4},
            mylang=eng]{my_style}


\begin{document}

\paragraph{Topics of the week} Compute with matrices:  

\begin{enumerate}
    \item do elimination and back substitution on square systems of linear equations, explain when and why this works or fails;
    \item define the inverse of a matrix, compute inverses of 2 × 2 matrices, characterize when the inverse exists (The Inverse Theorem), invert a product of matrices and the transpose of a matrix.
\end{enumerate}

\paragraph{CR decomposition} $A = CR$, where $C$ are \textit{the} independent columns of $A$.

\paragraph{Elementary operations}
\begin{itemize}
    \item $\boxed{A_i \gets A_i - c A_j}$ for $j \neq i$ and $c \in \RR$;
    \item $\boxed{A_i \gets c A_i}$ for any $c \neq 0$;
    \item $\boxed{A_i, A_j \gets A_j, A_i}$ for any $i \neq j$;
\end{itemize}

Can be represented by multiplying with a matrix on the left (see week 3).

\paragraph{Systems of linear equations} $Ax = b$. Solve by applying row operations to $A$ and $b$ until $A = I$.

\paragraph{Inverse matrix} $A^{-1}$ s.t. $A A^{-1} = A^{-1} A = I$. Unique if exists. How to find?

\begin{itemize}
    \item Solve $A (A^{-1} e_k) = e_k$ for all $e_k$ to find the $k$-th column of $A^{-1}$.
    \item Do elementary row operations on $I$ and $A$ simultaneously: $(A, I) \mapsto (I, A^{-1})$.
\end{itemize}

\paragraph{Determinant} The volume of the hypercube that is spanned on $(v_1,\dots,v_n)$. Properties:
\begin{enumerate}
    \item $V(e_1, \dots, e_n) = 1$.
    \item $V(a+b, \dots) = V(a, \dots) + V(b, \dots)$.
    \item $V(\alpha v, \dots) = \alpha V(\dots)$.
    \item $V(a, b, \dots) = -V(b, a, \dots)$.
\end{enumerate}

\paragraph{Inverse of $2 \times 2$} Try to derive Cramer's rule.

\paragraph{In-class exercises}

\begin{enumerate}
    \item Let $p : \RR \to \RR$ be a polynomial of degree at most $2$, i.e. $p(x) = ax^2+bx+c$.
    
    Find $a$, $b$ and $c$ s.t. $p(-1)=0$, $p(0)=2$ and $p(1) = 2$.

    $$
    \begin{pmatrix}
        (-1)^2 & -1 & 1 \\
        0^2 & 0 & 1 \\
        1^2 & 1 & 1
    \end{pmatrix} \begin{pmatrix} a \\ b \\ c \end{pmatrix} = 
    \begin{pmatrix}0 \\ 2 \\ 2\end{pmatrix}
    $$
    \item Assume that $Ax=b$ doesn't have an exact solution. Find $x$ that minimizes $\|Ax-b\|$.
\end{enumerate}

\end{document}
